Adversarial deep neural networks remove nonlinear batch effects from gene-expression data

### 2021-04-06 manuscript.tex
edits to manuscript including results

\begin{abstract}
	Gene-expression profiling enables researchers to quantify transcription levels in cells, thus providing insights into functional mechanisms of diseases and other biological processes.
	However, because of the high dimensionality of these data and the sensitivity of measuring equipment, expression data often contains unwanted confounding effects that can skew analyses.
	For example, collecting data in multiple runs can cause nontrivial differences in the data known as batch effects.
	In addition, covariates that are not of interest to the study may have strong effects on gene expression.
%	and there may be systemic effects when integrating multiple datasets.
	These confounding effects may be driven by higher-order interactions that are not removable using existing techniques that identify linear patterns.
	We created *Confounded* to remove these effects from expression data.
	[@Please modify the following sentence to better reflect the dual-objective aspect and the fact that it uses neural networks.@]
	Confounded is an adversarial variational autoencoder that removes confounding effects while minimizing the amount of change to the input data.
	We tested the model on artificially constructed data and three gene-expression datasets and compared against other batch-adjustment algorithms.
	Our software is available at \url{https://github.com/jdayton3/Confounded}.
\end{abstract}

@@

Gene-expression data can be used to advance our understanding of medicine and biology.
For example, gene-expression data have helped to discover conserved genetic modules \citep{stuart_gene-coexpression_2003}, to better understand the mechanisms of cardiovascular disease \citep{henriksen_application_2002}, to more accurately predict clinical outcomes in cancer patients \citep{veer_gene_2002}, and to discover effective drugs for treating specific diseases \citep{sirota_discovery_2011}.
Expression datasets are quite ``wide,'' typically containing tens of thousands of columns representing genes in the human transcriptome.
Because of the sensitive nature of technologies used to generate these data (i.e., expression levels may respond drastically to small environmental changes), expression data often contains unwanted confounding effects that can skew analyses.
Three examples of these include batch effects \label{it:batch}, known covariates, dataset-integration effects.

@@

Artificial neural networks, a category of machine-learning algorithms, are inspired by the way human brains function; input values pass through layers of linear and nonlinear functions; the final output values are measured against specific objectives, and the function layers are adjusted to bring the outputs closer to the objectives; this process is repeated until the outputs are sufficiently close to their targets \citep{schmidhuber_deep_2015}.
Neural networks have been applied broadly to gene-expression data. For example, they have been used to detect cancer and identify critical cancer genes \citep{danaee_deep_2016}, to impute gene-expression levels from the values of a few ``landmark genes'' \citep{chen_gene_2016}, to extract biologically relevant latent spaces in RNA-Seq data \citep{way_extracting_2017}, to reduce the dimensions of single-cell RNA-Seq data \citep{lin_using_2017}, to identify drug-repurposing targets \citep{aliper_deep_2016}, and to generate synthetic biomedical data as a way to preserve patient privacy \citep{beaulieu-jones_privacy-preserving_2017}.

Recent progress has been made in using artificial neural networks to correct batch effects in expression data \cite{shaham_removal_2017,shaham_batch_2018,upadhyay_removal_2019}.
However, these methods are designed for a limited range of scenarios:
1) when the input data contain only two batches,
2) the batches are sufficiently large (batch sizes smaller than 100 failed in our testing), and
3) the batches are balanced.
These requirements rarely hold in applied research; for example, the \textit{bladderbatch} dataset---which is used to test batch-adjustment techniques\cite{leek_bladderbatch_2017,leek_sva_2017}---has 5 batches, 57 samples, and 4 to 19 samples per batch.
Additionally, papers that have described these methods have not evaluated whether nonlinear patterns remain in expression data after batch correction.
To address these limitations, we implemented a deep neural-network architecture that 1) has a batch output layer that expands to fit
the number of batches, 2) works on relatively small batch sizes by sampling with replacement from the training data, and 3) has
no formulas that require batches to be equal sizes.
Furthermore, we tested our method using machine-learning classification algorithms to quantify the extent to which nonlinear batch effects remain after adjustment.

Autoencoders are a type of neural network that encodes input data in a smaller number of dimensions than the original data and then reconstructs the data in the original dimensions\citep{hinton_reducing_2006}.
This technique can be used to reduce noise and thus refine the data representation.[@Make sure this wording is legit. - Are you referencing denoising autoencoders, or are you saying that vanilla autoencoders reduce noise? Either way, we might need a reference. -You would know better than I. Please clarify this based on what is relevant to the paper and add a relevant reference.@]
Typically, autoencoders are trained using an objective function that seeks to ensure that the output is as similar as possible to the input \citep{hinton_reducing_2006}.
However, neural networks are adept at finding subtle patterns in data;
thus the objective function may prioritize replication of confounding patterns, such as batch effects, rather than the signals of interest \citep{ganin_domain-adversarial_2015,louizos_causal_2017-2}.
Recently, researchers have experimented with discouraging neural networks from learning such patterns by giving them two competing objective functions:
1) to maximize similarity between in the input and output data and
2) to ignore any patterns associated with known confounding variables \citep{ganin_domain-adversarial_2015,tzeng_deep_2014-2}.
For example, Louizos, et al. used this type of dual-objective function in combination with a variational autoencoder and successfully removed domain-based variability in credit scores, financial savings, and hospital admittance datasets \citet{louizos_variational_2015}.

@@

We implemented the variational autoencoder architecture \citep{louizos_variational_2015} described by \citet[Chapter 15]{geron_hands-machine_2017}.
This network has 2 hidden encoding layers and 2 decoding layers, each of size 500.
The code size is 20.
Each hidden layer is activated with the Exponential Linear Unit (ELU) function \citep{clevert_fast_2015} and trained with the Adam optimizer \citep{kingma_adam_2014} on reconstruction loss (sigmoid cross entropy) combined with latent loss (KL divergence \citep{kullback_information_1951}).

@@

\subsubsection{Training}

In all cases, we trained the network using the Adam Optimizer \citep{kingma_adam_2014} with a training rate of 0.0001 for 10,000 iterations on mini-batches of size 100.
In each iteration, we optimized on both $L_D$ and $L_{dual}$.
When optimizing $L_D$ (i.e. training the discriminator), we froze the autoencoder's weights, and vice versa.
We trained the network on a 2017 Dell XPS 15 9560 with an 8-core Intel i7-7700HQ central processing unit and 16 GB of random access memory.
For each dataset, training took roughly 30 minutes to complete, including time to load the input into memory and save the output to disk.

@@

We tested each method on four datasets that varied in size and the ype of data measured.

@@

@@

In a recent paper, \citet{shaham_removal_2017} used neural networks to remove batch effects.
Instead of constraining the autoencoder to remove batch effects based on a discriminator, these researchers trained their network to minimize maximum mean discrepancy (MMD) between batches in an embedded layer of their network.
We calculated MMD using the same formula to determine whether batches appear to come from the same distribution after adjustment.
We used the Gaussian kernel as implemented in \texttt{sklearn.metrics.pairwise.rbf\_kernel} \cite{pedregosa_scikit-learn_2011}.
In cases where there were more than two batches, we averaged all pairwise MMD values to calculate an overall MMD.

@@

%TODO: Update the algorithms listed here.
We used four classification algorithms from the scikit-learn (version 0.19.1) library \cite{pedregosa_scikit-learn_2011}: Naive Bayes \citep{maron_automatic_1961}, Random Forests \citep{tin_kam_ho_random_1995}, k-Nearest Neighbors \citep{fix_discriminatory_1951}, and Support Vector Machines (SVM) \citep{cortes_support-vector_1995} with a radial-basis kernel.
In most cases, we used the default hyperparameters, with the exceptions of using Random Forests with \texttt{n\_estimators=10} and using SVM with \texttt{kernel="rbf"}.
[@The defaults in version 0.19.1 of sklearn are n\_estimators=10 for RF and kernel="rbf" for SVM. Did you use these defaults or something different?@]

We calculated the average classification accuracy across four cross-validation folds, repeated three times.
We interpreted lower accuracy for batch classification to indicate that batch signal was removed more effectively.
We interpreted higher true-class accuracy to indicate that biologically relevant signal was retained during the batch-adjustment process.
Therefore, given output data from an ideal batch adjuster, batch classification accuracy would be no better than random chance, and true-class accuracy would be equal to what would be obtained for the unadjusted data.

@@

We would expect that after batch adjustment by an ideal adjuster, batch would no longer be detectable by any machine learning classifier.
Using the batch classification accuracy metric, Confounded seems to outperform other adjusters on larger datasets, whereas ComBat and Confounded seem to perform about the same on smaller datasets (see \figurename{} \ref{fig:batch}).
With both the bladderbatch and GSE37199 datasets, batch classification accuracy decreases well below baseline after batch adjustment with ComBat for all classifiers we tested (see Table \ref{tab:batch}).
Interestingly, batch accuracy also decreases drastically for the MNIST and TCGA datasets, but only for the Naive Bayes classifier.
This may be due to two factors: both ComBat and Naive Bayes use Bayesian methods, so ComBat may specifically remove the effects that Naive Bayes identifies; and Naive Bayes does not find patterns based on interactions between variables.
Although Naive Bayes is no longer able to identify confounding effects in the data after ComBat-adjustment, Random Forests (which does use interactions between variables) still has a very high accuracy for MNIST and an increased accuracy for TCGA.
In contrast, after adjustment by Confounded, the Random Forests algorithm's accuracy decreases more than with any other adjuster for both the MNIST and TCGA datasets.
This indicates that while ComBat's performance may work at least as well as Confounded for smaller expression datasets, Confounded may work better with larger datasets.

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/batch_accuracy.pdf}
	\caption[Batch classification accuracy]{\textbf{Batch classification accuracy} from 4-fold cross-validation repeated 3 times for several classifiers.
	Lower batch accuracy indicates that more batch-related signal has been removed and therefore indicates better performance.
	Confounded's performance is similar to ComBat's for the smaller datasets and is improved for the larger datasets.
	See also Table \ref{tab:batch}.}
	\label{fig:batch}
\end{figure}
\begin{table}
	\centering
	\input{tables/static/batch.tex}
	\caption[Batch classification accuracy]{
		\textbf{Batch classification accuracy} for several datasets and adjusters.
		The ideal batch adjuster would completely remove all signal due to batch and would therefore \textit{decrease} batch classification accuracy to around the baseline for all classifiers.
		See also \figurename{} \ref{fig:batch}.
	}
	\label{tab:batch}
\end{table}

With the larger datasets in particular, Confounded outperforms the other adjusters.
On the MNIST dataset, Random Forests is able to detect batch with perfect or near-perfect accuracy after adjustment with the scale adjuster and ComBat, but the highest batch classification accuracy after adjustment by Confounded is Naive Bayes, with an accuracy of 68.8\%.
With TCGA, both the scale adjuster and ComBat drastically increase Random Forests' batch classification accuracy from 87.6\% to 96.3\% and 97.1\% respectively, whereas Confounded decreases the accuracy to 8.8\%.

@@

With the smaller datasets, Confounded seems to keep true class information roughly as well as ComBat, (with Random Forests, Bladderbatch: 74.3 for ComBat\% and 72.1\% for Confounded, GSE37199: 60.4\% for ComBat and 69.0\% for Confounded; see \figurename{} \ref{fig:true_class} and Table \ref{tab:true_class}).
For the Bladderbatch dataset, true class accuracy is much lower after adjusting with any algorithm, indicating that cancer status and batch may not be independent.

With the larger datasets, Confounded's true class accuracy consistently decreases below the accuracy of other adjusters.
A look at the MNIST digits before and after adjustment (see \figurename{} \ref{fig:mnist}) shows that Confounded's output is often blurry, as is common with the output of variational autoencoders \cite{hou_deep_2016}.
With MNIST, Confounded's accuracy with Random Forests is still much higher than baseline (84.8\% vs. 11.4\%), but with TCGA, the accuracy decreases below baseline (66.5\% vs. 69.8\%) while the other adjusters' accuracies remain above baseline.
However, the particular set of parameters that we used in Confounded are likely not optimal for every dataset.
Additional tuning may improve the performance metrics.

@@

\textbf{Why should I use a neural network for batch adjustment?}
The process of measuring data typically leaves confounding effects, especially when the data or measurement process involves living things.
Imagine an experiment where two researchers measure the length of caterpillars.
If one researcher is left-eye dominant and the other is right-eye dominant, they may each see lengths as being a millimeter different.
If both researchers use the same wooden ruler at different relative humidities, the measurements may come out differently due to the wood swelling or due to the caterpillars reactions to humidity.
These same types of problems exist when measuring expression levels, but to more of an extreme---in addition to the target values and measuring equipment changing slightly due to systemic factors, the target values also change in response to each other.
Each of the 20,000 transcript levels may influence or be influenced by other transcript levels.
These cascading network-like effects are extremely likely to have nonlinear components, and our research shows that these nonlinear effects do indeed exist in gene expression data, thus rendering linear batch correction methods insufficient.
Therefore, some form of nonlinear adjustment must be used in order to correct for real-world confounding effects.
Rather than individually model each of infinitely many possible nonlinear interactions to see which represent confounders, we can use a neural network to both approximate and remove the confounders, since neural networks are proven to be universal function approximators \citep{csaji_approximation_2001}.

@@

\textbf{What limitations does Confounded have?}
\begin{enumerate*}[(a)]
	\item Confounded uses a variational autoencoder, which are known for often outputting a blurry version of the input data (as can be seen in \figurename{} \ref{fig:mnist}).
	However, recent work has identified modifications that may be made to the basic VAE structure to make output images sharper and more realistic \citep{hou_deep_2016}.
	Similar research with variational autoencoders and gene expression data may yield improved reconstruction losses and decrease the blurring effect.
	\item Confounded takes a long time to run in comparison with ComBat and other linear adjusters.
	Although we acknowledge this as a downfall of many types of machine learning and of neural network in particular, we believe that 30-60 minutes is a reasonable amount of time for a step that will be run only once per pipeline and that can greatly improve data quality.
	\item It can be difficult to identify the optimal network structure and parameter set for a neural network.
	Though this is the case for many applications of neural networks, we feel that Confounded's default structure worked well in our testing and that it will suffice for most batch correction applications.
	\item Neural networks usually perform better when given large amounts of data, and traditional batch datasets typically have very few samples.
	We did find that ComBat may outperform Confounded on many smaller, more traditional datasets, but that Confounded performs better on larger datasets.
	However, Confounded did perform reasonably well with the smaller datasets and appeared to avoid overfitting.
	In cases where ComBat is unable to completely remove confounding effects in a small dataset, Confounded may be a viable replacement method.
\end{enumerate*}

>>>>

\begin{abstract}
	Gene-expression profiling enables researchers to quantify transcription levels in cells, thus providing insights into functional mechanisms of diseases and other biological processes.
	However, because of the high dimensionality of these data and the sensitivity of measuring equipment, expression data often contains unwanted confounding effects that can skew analyses.
	For example, collecting data in multiple runs can cause nontrivial differences in the data known as batch effects.
	In addition, covariates that are not of interest to the study may have strong effects on gene expression.
%	and there may be systemic effects when integrating multiple datasets.
	These confounding effects may be driven by higher-order interactions that are not removable using existing techniques that identify linear patterns.
	We created *Confounded* to remove these effects from expression data.
	[@Please modify the following sentence to better reflect the dual-objective aspect and the fact that it uses neural networks.@] % duel objective meaning remove the effects and effectively replicate the origional expression levels/data? 
	Confounded is an adversarial variational autoencoder that removes confounding effects while minimizing the amount of change to the input data.
	We tested the model on artificially constructed data and three gene-expression datasets and compared against other batch-adjustment algorithms.
	Our software is available at \url{https://github.com/jdayton3/Confounded}.
	%I'm not sure if we need to put my repository on there since it's different than jonathan's now
\end{abstract}

@@

Gene-expression data can be used to advance our understanding of medicine and biology.
For example, gene-expression data have helped to discover conserved genetic modules \citep{stuart_gene-coexpression_2003}, to better understand the mechanisms of cardiovascular disease \citep{henriksen_application_2002}, to more accurately predict clinical outcomes in cancer patients \citep{veer_gene_2002}, and to discover effective drugs for treating specific diseases \citep{sirota_discovery_2011}.
Expression datasets are quite ``wide,'' typically containing tens of thousands of columns representing genes in the human transcriptome.
Because of the sensitive nature of technologies used to generate these data (i.e., expression levels may respond drastically to small environmental changes), expression data often contain unwanted confounding effects that can skew analyses. % removed an s from contains
Three examples of these include batch effects \label{it:batch}, known covariates, and dataset-integration effects. %included and
%kind of seems like we say for example/ and example many times.

@@

Artificial neural networks, a category of machine-learning algorithms, are inspired by the way human brains function; input values pass through layers of linear and nonlinear functions; the final output values are measured against specific objectives, and the function layers are adjusted to bring the outputs closer to the objectives; this process is repeated until the outputs are sufficiently close to their targets \citep{schmidhuber_deep_2015}.%There are a lot of semi colons here. Should we break up the sentence a little bit or reword?
Neural networks have been applied broadly to gene-expression data. For example, they have been used to detect cancer and identify critical cancer genes \citep{danaee_deep_2016}, to impute gene-expression levels from the values of a few ``landmark genes'' \citep{chen_gene_2016}, to extract biologically relevant latent spaces in RNA-Seq data \citep{way_extracting_2017}, to reduce the dimensions of single-cell RNA-Seq data \citep{lin_using_2017}, to identify drug-repurposing targets \citep{aliper_deep_2016}, and to generate synthetic biomedical data as a way to preserve patient privacy \citep{beaulieu-jones_privacy-preserving_2017}.

Progress has been made in using artificial neural networks to correct batch effects in expression data \cite{shaham_removal_2017,shaham_batch_2018,upadhyay_removal_2019}. % Was: Recent progress has been made ....
However, these methods are designed for a limited range of scenarios:
1) when the input data contain only two batches,
2) the batches are sufficiently large (batch sizes smaller than 100 failed in our testing), and
3) the batches are balanced.
These requirements rarely hold in applied research; for example, the \textit{bladderbatch} dataset---which is used to test batch-adjustment techniques\cite{leek_bladderbatch_2017,leek_sva_2017}---has 5 batches, 57 samples, and 4 to 19 samples per batch.
Additionally, papers that have described these methods have not evaluated whether nonlinear patterns remain in expression data after batch correction.
To address these limitations, we implemented a deep neural-network architecture that 1) has a batch output layer that expands to fit
the number of batches, 2) works on relatively small batch sizes by sampling with replacement from the training data, and 3) has
no formulas that require batches to be equal sizes.
Furthermore, we tested our method using machine-learning classification algorithms to quantify the extent to which nonlinear batch effects remain after adjustment.
% What is the difference between the first list and the second? Should we be consistant?

Autoencoders are a type of neural network that encodes input data in a smaller number of dimensions than the original data and then reconstructs the data in the original dimensions\citep{hinton_reducing_2006}.
This technique can be used to reduce noise and thus refine the data representation.[@Make sure this wording is legit. - Are you referencing denoising autoencoders, or are you saying that vanilla autoencoders reduce noise? Either way, we might need a reference. -You would know better than me. Please clarify this based on what is relevant to the paper and add a relevant reference.@]
Typically, autoencoders are trained using an objective function that seeks to ensure that the output is as similar as possible to the input \citep{hinton_reducing_2006}.
However, neural networks are adept at finding subtle patterns in data;
thus the objective function may prioritize replication of confounding patterns, such as batch effects, rather than the signals of interest \citep{ganin_domain-adversarial_2015,louizos_causal_2017-2}.
Researchers have experimented with discouraging neural networks from learning such patterns by giving them two competing objective functions: % was: Recently, researchers ....
1) to maximize similarity between in the input and output data and
2) to ignore any patterns associated with known confounding variables \citep{ganin_domain-adversarial_2015,tzeng_deep_2014-2}.
For example, Louizos, et al. used this type of dual-objective function in combination with a variational autoencoder and successfully removed domain-based variability in credit scores, financial savings, and hospital admittance datasets \citet{louizos_variational_2015}.

@@

%% ToDo: This is where we would have to change the default values in the confounded :
%% for DR Piccolo to Change : 
%%
%%
We implemented the variational autoencoder architecture \citep{louizos_variational_2015} described by \citet[Chapter 15]{geron_hands-machine_2017}.
This network has 2 hidden encoding layers and 2 decoding layers, each of size 500. % Is this sample size? I'm not sure where the 500 comes from
The code size is 200. %was: 20.
Each hidden layer is activated with the Exponential Linear Unit (ELU) function \citep{clevert_fast_2015} and trained with the Adam optimizer \citep{kingma_adam_2014} on reconstruction loss (sigmoid cross entropy) combined with latent loss (KL divergence \citep{kullback_information_1951}).

@@

\subsubsection{Training}
%% also another place to change the default values
In all cases, we trained the network using the Adam Optimizer \citep{kingma_adam_2014} with a training rate of 0.0001 for 10,000 iterations on mini-batches of size 256. % was: 100.
In each iteration, we optimized on both $L_D$ and $L_{dual}$.
When optimizing $L_D$ (i.e. training the discriminator), we froze the autoencoder's weights, and vice versa.
We trained the network on a 2017 Dell XPS 15 9560 with an 8-core Intel i7-7700HQ central processing unit and 16 GB of random access memory.
For each dataset, training took roughly 30 minutes to complete, including time to load the input into memory and save the output to disk. % We should probably re-time it to make sure this is still accurate

@@

We tested each method on four datasets that varied in size and the type of data measured.

@@

%% I keep forgetting whether or not we are leaving the MNIST dataset and just not doing anything, or if we are deleting it out. 
%% maybe put in the simulated expression data and information here

%\subsection{Simulated Expression}
%We simulated a balanced two class two batch dataset that contains 200 samples of 1000 features each that allowed us to simulate a large dataset to optimize the parameters of the confounded method.
%Initially, we used make_classification from the sklearn classification library \url{}, but combat always outperformed confounded because the clustering made by the make_classification algorithm made it possible for linear models to accurately classify the true classes.
%As we moved away from clustering and added nonlinear patterns to a randomly generated normal distribution set, we saw confounded begin to perform better than combat. 
%When the true class and batch signals are nonlinear and small, combat is not as able to acuratly classify the classes or remove the batches, but confounded is able to identify the classes and effectively remove the batches.

@@

In an earlier paper, \citet{shaham_removal_2017} used neural networks to remove batch effects.
%WAS: In a recent paper, \citet{shaham_removal_2017} used neural networks to remove batch effects.
Instead of constraining the autoencoder to remove batch effects based on a discriminator, these researchers trained their network to minimize maximum mean discrepancy (MMD) between batches in an embedded layer of their network.
We calculated MMD using the same formula to determine whether batches appear to come from the same distribution after adjustment.
We used the Gaussian kernel as implemented in \texttt{sklearn.metrics.pairwise.rbf\_kernel} \cite{pedregosa_scikit-learn_2011}.
In cases where there were more than two batches, we averaged all pairwise MMD values to calculate an overall MMD.

@@

%TODO: Update the algorithms listed here.
We used four classification algorithms from the scikit-learn (version 0.19.1) library \cite{pedregosa_scikit-learn_2011}: Random Forests \citep{tin_kam_ho_random_1995}, k-Nearest Neighbors \citep{fix_discriminatory_1951}, and Support Vector Machines (SVM) \citep{cortes_support-vector_1995} with a radial-basis kernel.
%We used four classification algorithms from the scikit-learn (version 0.19.1) library \cite{pedregosa_scikit-learn_2011}: Naive Bayes \citep{maron_automatic_1961}, Random Forests \citep{tin_kam_ho_random_1995}, k-Nearest Neighbors \citep{fix_discriminatory_1951}, and Support Vector Machines (SVM) \citep{cortes_support-vector_1995} with a radial-basis kernel.
In most cases, we used the default hyperparameters, with the exceptions of using Random Forests with \texttt{n\_estimators=10} and using SVM with \texttt{kernel="rbf"}.
[@The defaults in version 0.19.1 of sklearn are n\_estimators=10 for RF and kernel="rbf" for SVM. Did you use these defaults or something different?@] %we used those parameters

We calculated the average classification accuracy across three cross-validation folds, repeated ten times.
%We calculated the average classification accuracy across four cross-validation folds, repeated three times.
We interpreted lower accuracy for batch classification to indicate that batch signal was removed more effectively.
We interpreted higher true-class accuracy to indicate that biologically relevant signal was retained during the batch-adjustment process.
Therefore, given output data from an ideal batch adjuster, batch classification accuracy would be no better than random chance, and true-class accuracy would be equal to what would be obtained for the unadjusted data.

@@

We would expect that after batch adjustment by an ideal adjuster, batch would no longer be detectable by any machine learning classifier.
Using the batch classification accuracy metric, Confounded seems to outperform other adjusters on larger datasets, whereas ComBat and Confounded seem to perform about the same on smaller datasets (see \figurename{} \ref{fig:batch}).
With both the bladderbatch and GSE37199 datasets, batch classification accuracy decreases well below baseline after batch adjustment with ComBat for all classifiers we tested (see Table \ref{tab:batch}).
%Interestingly, batch accuracy also decreases drastically for the MNIST and TCGA datasets, but only for the Naive Bayes classifier.
%This may be due to two factors: both ComBat and Naive Bayes use Bayesian methods, so ComBat may specifically remove the effects that Naive Bayes identifies; and Naive Bayes does not find patterns based on interactions between variables.
%Although Naive Bayes is no longer able to identify confounding effects in the data after ComBat-adjustment, Random Forests (which does use interactions between variables) still has a very high accuracy for MNIST and an increased accuracy for TCGA.
In contrast, after adjustment by Confounded, the Random Forests algorithm's accuracy decreases more than with any other adjuster for both the MNIST and TCGA datasets.
This indicates that while ComBat's performance may work at least as well as Confounded for smaller expression datasets, Confounded may work better with larger datasets.

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/batch_accuracy.pdf}
	\caption[Batch classification accuracy]{\textbf{Batch classification accuracy} from 3-fold cross-validation repeated 10 times for several classifiers.
	Lower batch accuracy indicates that more batch-related signal has been removed and therefore indicates better performance.
	Confounded's performance is similar to ComBat's for the smaller datasets and is improved for the larger datasets.
	See also Table \ref{tab:batch}.}
	\label{fig:batch}
\end{figure}
\begin{table}
	\centering
	\input{tables/static/batch.tex}
	\caption[Batch classification accuracy]{
		\textbf{Batch classification accuracy} for several datasets and adjusters.
		The ideal batch adjuster would completely remove all signal due to batch and would therefore \textit{decrease} batch classification accuracy to around the baseline for all classifiers.
		See also \figurename{} \ref{fig:batch}.
	}
	\label{tab:batch}
\end{table}

With the larger datasets in particular, Confounded outperforms the other adjusters.
On the MNIST dataset, Random Forests is able to detect batch with perfect or near-perfect accuracy after adjustment with the scale adjuster and ComBat, but the highest batch classification accuracy after adjustment by Confounded is Naive Bayes, with an accuracy of 68.8\%.
With TCGA, both the scale adjuster and ComBat drastically increase Random Forests' batch classification accuracy from 95.0\% to 98.5\% and 99.9\% respectively, whereas Confounded decreases the accuracy to 77.8\%.
%With TCGA, both the scale adjuster and ComBat drastically increase Random Forests' batch classification accuracy from 87.6\% to 96.3\% and 97.1\% respectively, whereas Confounded decreases the accuracy to 8.8\%.

@@

%What is the true class information for Bladderbatch?
With the smaller datasets, Confounded seems to keep true class information roughly as well as ComBat, (with Random Forests, Bladderbatch: 74.3 for ComBat\% and 72.1\% for Confounded, GSE37199: 68.7\% for ComBat and 68.8\% for Confounded; see \figurename{} \ref{fig:true_class} and Table \ref{tab:true_class}).
%With the smaller datasets, Confounded seems to keep true class information roughly as well as ComBat, (with Random Forests, Bladderbatch: 74.3 for ComBat\% and 72.1\% for Confounded, GSE37199: 60.4\% for ComBat and 69.0\% for Confounded; see \figurename{} \ref{fig:true_class} and Table \ref{tab:true_class}).
For the Bladderbatch dataset, true class accuracy is much lower after adjusting with any algorithm, indicating that cancer status and batch may not be independent.

With the larger datasets, Confounded's true class accuracy consistently decreases below the accuracy of other adjusters.
%%This isn't accurate anymore. Our fine tuning made it better. 
%%This may be a good place to include simulated expression results. 
The Confounded's accuracy with Random Forest on our simulated dataset is still much higher than the baseline(70.5\% vs. 50.0\%). %Not sure if that's the right wording
A look at the TCGA data before and after adjustment shows that even with larger datasets, the accuracy of Confounded never decreases below the baseline (72.6\% vs. 69.8\%). 
% compare TCGA datasets here to show that the size of datasets affects the performance. 
%To compare the effect that the size of the data has on Confounded in relation to the other methods, we created two sub-datasets from the TCGA data, a medium and a small and compared their performance. On Random Forest, the ability for machine learning to identify smaples with TP53_Mutated genes as the data size increased changed in Scaled: 71.2 \% to 71.9\% to 80.1\%, in Combat: 71.0\% to 75.2\% to 81.0\%, and Confounded: 67.9\% to 71.0\% to 72.6\%.
%The ability for machine learning classification algorithms to detect the CancerType also increased with increase in size of the dataset. The Random Forest accuracy increased in each adjuster with Scaled: 7.2\% to 89.0\% to 98.5\%, in Combat: 96.3\% to 99.9\% to 99.9\%, and Confounded: 65.8\% to 66.5\% to 78.0\%.
%Though the ability for Confounded to replicate the origional expression data is a little lower than that of the other adjusters, the ability for the Confounded to remove the patterns that allow machine learning to detect CancerType is sgnificantly better than that of the other adjusters which have very high, and sometimes higher accuracy after adjustment than before any kind of adjustment is made. See table? 
%A look at the MNIST digits before and after adjustment (see \figurename{} \ref{fig:mnist}) shows that Confounded's output is often blurry, as is common with the output of variational autoencoders \cite{hou_deep_2016}.
%With MNIST, Confounded's accuracy with Random Forests is still much higher than baseline (84.8\% vs. 11.4\%), but with TCGA, the accuracy decreases below baseline (66.5\% vs. 69.8\%) while the other adjusters' accuracies remain above baseline.
%However, the particular set of parameters that we used in Confounded are likely not optimal for every dataset.
Additional tuning may improve the performance metrics.

@@

\textbf{Why should I use a neural network for batch adjustment?}
The process of measuring data typically leaves confounding effects, especially when the data or measurement process involves living things.
Imagine an experiment where two researchers measure the length of caterpillars.
If one researcher is left-eye dominant and the other is right-eye dominant, they may each see lengths as being a millimeter different.
If both researchers use the same wooden ruler at different relative humidities, the measurements may come out differently due to the wood swelling or due to the caterpillars reactions to humidity.
These same types of problems exist when measuring expression levels, but to more of an extreme---in addition to the target values and measuring equipment changing slightly due to systemic factors, the target values also change in response to each other.
Each of the 20,000 transcript levels may influence or be influenced by other transcript levels.
These cascading network-like effects are extremely likely to have nonlinear components, and our research shows that these nonlinear effects do indeed exist in gene expression data, thus rendering linear batch correction methods insufficient.
Therefore, some form of nonlinear adjustment must be used in order to correct for real-world confounding effects.
%It may just be me, but this senstence is confusing because of the length and the number of commas
Rather than individually model each of infinitely many possible nonlinear interactions to see which represent confounders, we can use a neural network to both approximate and remove the confounders, since neural networks are proven to be universal function approximators \citep{csaji_approximation_2001}.

@@

\textbf{What limitations does Confounded have?}
\begin{enumerate*}[(a)]
	\item Confounded uses a variational autoencoder, which are known for often outputting a blurry version of the input data (as can be seen in \figurename{} \ref{fig:mnist}).
	However, work in 2016 has identified modifications that may be made to the basic VAE structure to make output images sharper and more realistic \citep{hou_deep_2016}
%	However, recent work has identified modifications that may be made to the basic VAE structure to make output images sharper and more realistic \citep{hou_deep_2016}.
	Similar research with variational autoencoders and gene expression data may yield improved reconstruction losses and decrease the blurring effect.
	\item Confounded takes a long time to run in comparison with ComBat and other linear adjusters.
	Although we acknowledge this as a downfall of many types of machine learning and of neural networks in particular, we believe that 30-60 minutes is a reasonable amount of time for a step that will be run only once per pipeline and that can greatly improve data quality.
	\item It can be difficult to identify the optimal network structure and parameter set for a neural network.
	Though this is the case for many applications of neural networks, we feel that Confounded's default structure worked well in our testing and that it will suffice for most batch correction applications.
	% I wrote that in simulated expression that we optimized the parameters, so we could say that we optimized parameters that worked well across several different datasets and should be good for most applications, but can further be optimized to perform better on individual datasets.
	\item Neural networks usually perform better when given large amounts of data, and traditional batch datasets typically have very few samples.
	We did find that ComBat may outperform Confounded on many smaller, more traditional datasets, but that Confounded performs better on larger datasets.
	However, Confounded did perform reasonably well with the smaller datasets and appeared to avoid overfitting.
	In cases where ComBat is unable to completely remove confounding effects in a small dataset, Confounded may be a viable replacement method.
\end{enumerate*}
